---
title: "Cell Type Composition"
author: "Zhaoheng Li, Wei Sun"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: journal
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: false
    number_sections: false
  df_print: paged
---

> condition by time point

> only consider time points in KO village samples

# libraries

```{r load_libraries, warning = FALSE, message = FALSE}
library(Seurat)
library(Matrix)
library(dplyr)
library(tidyr)
library(ggplot2)
library(tibble)
library(cowplot)  
library(patchwork)
library(stringr)
library(pheatmap)
library(reshape2)
library(readxl)
library(stringr)
library(DESeq2)
library(gridExtra)
library(ggrepel)
library(ggpubr)
library(tidyverse)

theme_set(theme_classic())


data_dir = "/home/zli4/MSK_DEC24/Analysis"

out_dir = "/fh/fast/sun_w/zhaoheng_li/MSK_DEC24/Analysis/outs/cell_type_composition"

set.seed(022520)
```


```{r}
lm_plot=function(mod){
  summary_model= summary(mod)
  coefficients= summary_model$coefficients
  coeff_df= as.data.frame(coefficients)
  coeff_df$Predictor= rownames(coeff_df)

  significant_predictors= coeff_df %>%
    filter(Predictor != "(Intercept)" & `Pr(>|t|)` < 0.05)
  

  conf_int= confint(mod)
  conf_df= as.data.frame(conf_int)
  conf_df$Predictor= rownames(conf_df)
  
  # Merge with the significant predictors
  plot_data= merge(significant_predictors, conf_df, by = "Predictor")
  names(plot_data)= c("Predictor", "Estimate", "Std.Error", "t.value", "p.value", "CI.Lower", "CI.Upper")
  
  plot_data$Predictor = gsub(paste(c("genotype"), collapse = "|"), "", plot_data$Predictor)
  fig = ggplot(plot_data, aes(x = reorder(Predictor, Estimate), y = Estimate)) +
    geom_point() +
    geom_errorbar(aes(ymin = CI.Lower, ymax = CI.Upper), width = 0.2) +
    coord_flip() +  # Flip the coordinates for better readability
    theme_minimal() +
    labs(
      title = "Estimates and Confidence Intervals for Significant Predictors",
      x = "",
      y = "Estimate"
    )
  return(fig)
}

lm_plot_exp=function(mod){
  summary_model= summary(mod)
  coefficients= summary_model$coefficients
  coeff_df= as.data.frame(coefficients)
  coeff_df$Predictor= rownames(coeff_df)

  significant_predictors= coeff_df %>%
    filter(Predictor != "(Intercept)" & `Pr(>|t|)` < 0.05)
  

  conf_int= confint(mod)
  conf_df= as.data.frame(conf_int)
  conf_df$Predictor= rownames(conf_df)
  
  # Merge with the significant predictors
  plot_data= merge(significant_predictors, conf_df, by = "Predictor")
  names(plot_data)= c("Predictor", "Estimate", "Std.Error", "t.value", "p.value", "CI.Lower", "CI.Upper")
  
  plot_data$Predictor = gsub(paste(c("genotype"), collapse = "|"), "", plot_data$Predictor)
  fig = ggplot(plot_data, aes(x = reorder(Predictor, Estimate), y = exp(Estimate))) +
    geom_point() +
    geom_errorbar(aes(ymin = exp(CI.Lower), ymax = exp(CI.Upper)), width = 0.2) +
    coord_flip() +  # Flip the coordinates for better readability
    theme_minimal() +
    labs(
      title = "Estimates and Confidence Intervals for Significant Predictors",
      x = "",
      y = "Estimate"
    )
  return(fig)
}
```




# load data

```{r}
integrated = readRDS(file.path(data_dir,"integrated-rpca_BDF-ref_withMeta.RDS"))
```

```{r}


# remove cells without cell type annotations
keep.cells= rownames(integrated@meta.data)[
  !is.na(integrated@meta.data$celltype)
]

integrated= subset(
  integrated,
  cells = keep.cells
)
integrated$celltype = as.factor(integrated$celltype)
summary(integrated$celltype)
```

# order cell types

```{r}
celltype_orders = c(
  "ESC",
  "ESC_DE",
  "DE",
  "PFG",
  "PGT",
  "PP",
  "EnP",
  "SC-EC",
  "SC-alpha",
  "SC-beta",
  "SC-delta",
  "Ductal",
  "Mesenchymal_Stromal",
  "Stromal",
  "Endothelial",
  "Liver"
)

integrated$celltype = factor(integrated$celltype,
                              levels =celltype_orders ,
                              ordered = TRUE)

tol_okabe16 <- c(
  "#4477AA", "#EE6677", "#228833", "#CCBB44",
  "#66CCEE", "#AA3377", "#BBBBBB", "#009E73",
  "#E69F00", "#56B4E9", "#0072B2", "#D55E00",
  "#CC79A7", "#117733", "#882255", "#DDCC77"
)

celltype_palette = setNames(
  tol_okabe16,
  celltype_orders
)
saveRDS(celltype_palette,"/fh/fast/sun_w/zhaoheng_li/MSK_DEC24/Analysis/outs/celltype_palette.RDS")
celltype_palette = readRDS("/fh/fast/sun_w/zhaoheng_li/MSK_DEC24/Analysis/outs/celltype_palette.RDS")
```

# filter by time points

```{r}
obj = integrated%>%subset(subset = time_point%in%c("D-1","D3","D7","D11","D18"))

obj$time_point = droplevels(obj$time_point)
summary(obj$time_point)
```

```{r}

meta_df = obj[[]]

cell_counts = meta_df %>%
  group_by(time_point, celltype, feature.4) %>%
  summarise(
    n_cells = n(),
    .groups = "drop"
  )


head(cell_counts)
write.table(
  cell_counts,
  file      = file.path(out_dir,"cell_counts.tsv"),
  sep       = "\t",
  quote     = FALSE,
  row.names = FALSE
)
```

## sample by time point

```{r}
table(obj$sample_name,obj$time_point)
```


## genotype by time point

```{r}
table(obj$genotype,obj$time_point)

```

## cell type by time point

```{r}
table(obj$celltype,obj$time_point)

```


# cell-type proportions by time_point

```{r}
meta = obj[[]]
```



```{r,fig.width=8,fig.height=6}


# ignore time points unique to external WT samples
common_times = rownames(table(meta$time_point,meta$source))[rowSums(table(meta$time_point,meta$source)!=0)==2] 


prop_by_time = meta %>%
  filter(time_point %in% common_times) %>%
  group_by(time_point, celltype) %>%
  summarise(n = n(), .groups = "drop_last") %>%
  mutate(prop = n / sum(n)) %>%         # within each time_point
  ungroup()

fig1= prop_by_time%>%ggplot(aes(x = time_point, y = prop, fill = celltype)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent_format(1)) +
  scale_fill_manual(
    values = celltype_palette,   
    drop   = TRUE                
  )+
  labs(
    title = "cell-type proportions by time point",
    x     = "",
    y     = "",
    fill  = ""
  ) +
  guides(fill = guide_legend(nrow = 3, byrow = TRUE)) +
  theme_classic() +
  theme(
    legend.position   = "top",           # ← move legend to top
    legend.direction  = "horizontal",    # ← ensure it’s laid out horizontally
    axis.text.x       = element_text(angle = 45, hjust = 1)
  )
fig1
ggsave(
  filename = file.path(out_dir,"celltype_prop_time.png"),
  plot     = fig1,
  width    = 8,
  height   = 4,
  units    = "in",
  dpi      = 300
)

```

#  cell-type proportions by genotype

```{r,fig.width=8,fig.height=6}


prop_by_genotype = meta %>%
  group_by(genotype, celltype) %>%
  summarise(n = n(), .groups = "drop_last") %>%
  mutate(prop = n / sum(n)) %>%         # within each genotype
  ungroup()

fig2 = prop_by_genotype %>%
  ggplot(aes(x = genotype, y = prop, fill = celltype)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent_format(1)) +
  scale_fill_manual(
    values = celltype_palette,   
    drop   = TRUE                
  )+
  labs(
    title = "cell-type proportions by genotype",
    x     = "",
    y     = "",
    fill  = ""
  ) +
  guides(fill = guide_legend(nrow = 3, byrow = TRUE)) +
  theme_classic() +
  theme(
    legend.position   = "top",           # ← move legend to top
    legend.direction  = "horizontal",    # ← ensure it’s laid out horizontally
    axis.text.x       = element_text(angle = 45, hjust = 1)
  )

ggsave(
  filename = file.path(out_dir,"celltype_prop_genotype.png"),
  plot     = fig2,
  width    = 8,
  height   = 4,
  units    = "in",
  dpi      = 300
)
fig2
```




# split seurat object by time points


```{r}
obj_list = SplitObject(obj, split.by = "time_point")
obj_list
```




# cell-type proportions by genotype (conditioned on time point)


```{r, warning = FALSE, message=FALSE}
prop_lst = lapply(c("D3","D7","D11","D18"),function(tp){
  df = obj_list[[tp]][[]]
  df %>% group_by(genotype, celltype) %>%
  summarise(count = n()) %>%
  mutate(proportion = count / sum(count)) %>%
  arrange(genotype, desc(proportion))
})
names(prop_lst) = c("D3","D7","D11","D18")
```




```{r,fig.width=8,fig.height=6}

fig_list = lapply( c("D3","D7","D11","D18"),function(tp){
  df = prop_lst[[tp]] 
  df %>%
    ggplot(aes(x = genotype, y = proportion, fill = celltype)) +
    geom_col() +
    scale_y_continuous(labels = scales::percent_format(1)) +
    scale_fill_manual(
      values = celltype_palette,   
      drop   = TRUE      
    )+
    labs(
      title = tp,
      x     = "",
      y     = "cell-type proportions",
      fill  = ""
    ) +
    guides(fill = guide_legend(nrow = 3, byrow = TRUE)) +
    theme_classic() +
    theme(
      legend.position   = "top",           # ← move legend to top
      legend.direction  = "horizontal",    # ← ensure it’s laid out horizontally
      axis.text.x       = element_text(angle = 45, hjust = 1)
    )
})
names(fig_list) = c("D3","D7","D11","D18")

for(tp in names(fig_list)){
  ggsave(file.path(out_dir,paste0("genotype_celltype_proportion/",tp,".png",sep='')),
         plot     = fig_list[[tp]],
  width    = 8,
  height   = 6,
  units    = "in",
  dpi      = 300
  )
}

fig_list
```

# genotype-wise similarity using cell type compositions (conditioned on time point)

> using centering log ratio transformation

```{r, fig.width=8,fig.height=8, warning = FALSE, message= FALSE}
fig_list = list()
for(tp in names(prop_lst)){
  
  df = prop_lst[[tp]]
  df$celltype = as.factor(df$celltype)
  levels(df$celltype)=gsub("_","-",levels(df$celltype)) # rename celltype levels
  
  df = df%>%
    group_by(genotype)%>%
    mutate(geom_mean = exp(mean(log(proportion))))%>%
    mutate(clr = log(proportion/geom_mean))
  
  df$celltype = droplevels(df$celltype)
  print(unique(df$celltype))
  
  clr = df %>%
    select(genotype, celltype, clr) %>%
    tidyr::spread(key = celltype, value = clr) %>%
    tibble::column_to_rownames(var = "genotype") %>%
    as.matrix()
  
  clr[is.na(clr)] = 2*min(clr,na.rm=T) # deal with NA values
  
  clr_dist = as.matrix(dist(clr, method = "euclidean"))
  hc = hclust(as.dist(clr_dist))
  d = mean(  clr_dist)
  fig = pheatmap::pheatmap(exp(-clr_dist^2/(2*d)),
                           clustering_method = 'ward.D',
                           cluster_rows = hc,cluster_cols =hc,main = tp,fontsize = 5)
  fig_list[[tp]] = fig[[4]]
}
g = gridExtra::grid.arrange(gridExtra::arrangeGrob(grobs= fig_list,ncol=2))
ggsave(file.path(out_dir,"geno_similarity_cell-type-composition.png"),g)
g
```


# create pseudobulk data



```{r make-pseudobulk, message = FALSE, warning=FALSE}
f_pseudobulk_list = file.path(out_dir,"ct_composition_pseudobulk_list.RDS")

if(file.exists(f_pseudobulk_list)){
  pseudo_obj_list = readRDS(f_pseudobulk_list)
}else{
  pseudo_obj_list = list()
  for(tp in c("D-1","D3","D7","D11","D18")){
    print(paste0("working on ", tp))
    obj = obj_list[[tp]]
    obj$background = as.factor(obj$background)
    obj$genotype = as.factor(obj$genotype)
    levels(obj$celltype)=gsub("_","-",levels(obj$celltype)) # rename celltype levels
    obj$celltype = droplevels(obj$celltype)
    obj$pseudo.sample.id = as.factor(paste(obj$sample_name,obj$background,obj$genotype,obj$feature.4,sep='_')) # handle cell type names
    pseudo_obj= AggregateExpression(obj, assays = "RNA", return.seurat = T,
                                        group.by = c("sample_name","background","genotype", "feature.4"))
    obj$pseudo.sample.id =  gsub("-","_",obj$pseudo.sample.id)
    pseudo_obj$pseudo.sample.id = gsub("-","_", colnames(pseudo_obj))
    pseudo_meta = pseudo_obj[[]]
    df = obj[[]] %>% 
      group_by(pseudo.sample.id, celltype) %>%
    summarise(count = n(),.groups = "drop") %>%
    tidyr::complete(pseudo.sample.id, celltype, fill = list(count = 0))%>%
       group_by(pseudo.sample.id) %>%
    mutate(proportion = count / sum(count)) 
    
    df_wide = df%>%tidyr::pivot_wider(names_from = celltype, values_from = c(count,proportion))
    pseudo_meta  = pseudo_meta %>%left_join(df_wide)
    rownames(pseudo_meta) = pseudo_meta$orig.ident
    
    n.df=obj[[]]%>%dplyr::count(pseudo.sample.id)
    pseudo_meta = pseudo_meta %>%left_join(n.df)
    rownames(pseudo_meta) = pseudo_meta$orig.ident
    
    pseudo_obj@meta.data =  pseudo_meta 
    
    pseudo_obj$source =as.factor(ifelse(pseudo_obj$sample_name%in%c("A","B","C","D","E","F"),"external_WT", "KO_village" ))
    
    pseudo_obj_list[[tp]] = pseudo_obj
  }
  saveRDS(pseudo_obj_list,f_pseudobulk_list)
}
```


```{r}
names(celltype_palette) = gsub("_","-",names(celltype_palette))
```

## check sample compositions for each time point

```{r}
for(tp in c("D-1","D3","D7","D11","D18")){
  print(tp)
  df = pseudo_obj_list[[tp]][[]]
  print(summary(as.factor(df$sample_name)))
}

```


# remove pseudobulk samples with < 30 cells

```{r}
min_cells = 30
```


```{r}

pseudo_obj_list = lapply(pseudo_obj_list,function(pseudo_obj){
  pseudo_obj%>%subset(subset = n>=min_cells)
})
pseudo_obj_list
```



# barplot for pseudobulk cell type proportions

```{r proportion-barplots, fig.width = 12,fig.height=10}
barplot_list = list()
for(tp in c("D3","D7","D11","D18")){
  pseudo_obj = pseudo_obj_list[[tp]]
  df = pseudo_obj[[]]
  df_long= df %>%
  tidyr::pivot_longer(cols = starts_with("proportion_"), 
               names_to = "celltype", 
               values_to = "Proportion")
  df_long$celltype = gsub("proportion_","",df_long$celltype) 
  df_long = df_long %>%
  mutate(
    celltype = factor(celltype, levels = gsub("_","-",celltype_orders),ordered = TRUE)
  )
  
  fig = ggplot(df_long, aes(x = pseudo.sample.id, y = Proportion, fill = celltype)) +
    geom_col(position = "stack") +
    scale_fill_manual(
      values =celltype_palette,   
      drop   = TRUE                
    )+
    facet_wrap(~ genotype, 
               nrow = 2, 
               scales = "free_x", 
               strip.position = "top") +
    ylab("cell type proportion") +
    xlab("pseudobulk samples") +
    labs(
      title = tp,
      x     = "pseudobulk samples",
      y     = "cell type proportion",
      fill  = ""
    ) +
    guides(fill = guide_legend(nrow = 2, byrow = TRUE)) +
    theme_pubr() +
    theme(
      strip.text = element_text(angle = 90, size = 8, hjust = 0, face = "bold"),
      strip.background = element_rect(colour = "black", fill = "grey95", linewidth = 0),
      axis.text.x = element_blank(),
      axis.line = element_blank(),
      axis.ticks = element_blank()
    )

  barplot_list[[tp]] = fig
}

for(tp in names(barplot_list)){
  ggsave(file.path(out_dir,paste0("pseudobulk_celltype_proportion/",tp,".png",sep='')),
         plot     = barplot_list[[tp]],
  width    = 12,
  height   = 10,
  units    = "in",
  dpi      = 300
  )
}


barplot_list
```






# check distribution of cell type log proportions


```{r}

for( tp in c("D3","D7","D11","D18")){
  obj = pseudo_obj_list[[tp]]
  df =  obj[[]]
  ct_response =  gsub("count_","",grep("^count_", names(df), value = TRUE))
  n = nrow(df)
  # create log proportions
  for(ct in ct_response){
    df[,paste("proportion0",ct,sep="_")] = df[,paste("proportion",ct,sep="_")]*(n-1)/n+0.5/n # transformation used by DR_data
    df[[paste("log_prop", ct,sep="_")]] = log(df[,paste("proportion0",ct,sep="_")] )
    print(paste(ct, " # Inf: ", sum(is.infinite(df[[paste("log_prop", ct,sep="_")]] )),sep=''))
  }
  
  for(c in grep("^log_prop_", colnames(df), value = TRUE)){
    hist(as.vector(df[c])[[1]],main = paste(tp, c,sep=" : "),xlab = "log proportion")
  }
}


```

# run linear model and plot 95% confidence intervals of significant predictors


```{r linear-model}
fig_list =list()
coef_list = list()
p_list = list()
for( tp in c("D3","D7","D11","D18")){
  df =  pseudo_obj_list[[tp]][[]]
  ct_response = gsub("count_","",grep("^count_", names(df), value = TRUE))
  n = nrow(df)
  # create log proportions
  for(ct in ct_response){
    df[,paste("proportion0",ct,sep="_")] = df[,paste("proportion",ct,sep="_")]*(n-1)/n+0.5/n # transformation used by DR_data
    df[[paste("log_prop", ct,sep="_")]] = log(df[,paste("proportion0",ct,sep="_")] )
  }
  
  mod_tp = list()
  coef_mtx = list()
  p_mtx = list()
  for(ct in ct_response){
    # run regression
    if(max(df[,paste("proportion",ct,sep="_")])>0.03){
      df$genotype = as.factor(df$genotype)
      df$genotype = relevel(df$genotype, ref = "WT")
      v = as.symbol(paste("log_prop", ct,sep="_"))
      mod = eval(bquote(lm(.(v)~source+background+genotype,data=df)))
      mod_tp[[ct]] = mod
      coef_mtx[[ct]] = summary(mod)$coefficients[grepl("genotype", names(summary(mod)$coefficients[,4])),1]
      p_mtx[[ct]] = summary(mod)$coefficients[grepl("genotype", names(summary(mod)$coefficients[,4])),4]
    }
  }
  coef_mtx = do.call(cbind,coef_mtx);rownames(coef_mtx) = gsub("genotype","",rownames(coef_mtx) )
  p_mtx = do.call(cbind,p_mtx);rownames(p_mtx) = gsub("genotype","",rownames(p_mtx) )
  coef_list[[tp]] = coef_mtx
  p_list[[tp]] = p_mtx
  
  # visualize results
  for(ct in names(mod_tp)){
    if(any(p_mtx[,ct]<=0.05)){
          mod = mod_tp[[ct]]
      fig = lm_plot(mod)+
                   ggtitle(paste(tp, ":", ct,sep=''))+
        theme(plot.title = element_text(size = 10, face = "bold"))
      fig_list[[paste(tp,ct,sep="_")]] = fig
    }
  }
}

```




> only cell types with at least one genotype as a significant predictors are shown

```{r coef-plot,fig.height = 15,fig.width=15, message = FALSE, warning = FALSE}

coef_fig = gridExtra::grid.arrange(grobs = fig_list,ncol = 6,top = "Significant Coefficients on cell-type log(proportion)")
print(coef_fig)
ggsave(file.path(out_dir,"log_proportion_lm_coef.png"),coef_fig,width = 15,height=15)

```

# heatmaps of estimated coefficients for genotypes

> values are truncated at $\pm 3$ for visualization purpose

```{r}
meta%>%filter(time_point=='D3')%>%group_by(celltype)%>%summarise(n=n())
```



```{r coef-htmap}

htmap_list = list()
for(tp in names(coef_list)){
  coef_mtx =coef_list[[tp]]
  coef_mtx[coef_mtx>3] = 3;coef_mtx[coef_mtx=3] = -3
  htmap = pheatmap::pheatmap(coef_mtx,cluster_cols = F,fontsize = 5,
                      color = colorRampPalette(c("blue", "white", "red"))(50),
                      breaks = seq(min(coef_mtx,na.rm=T), max(coef_mtx,na.rm=T), length.out = 51),
                     main = tp)
  htmap_list[[tp]] = htmap
}
names(htmap_list) = names(coef_list)

for(tp in names(htmap_list)){
  png(file.path(out_dir,paste0("coefs_htmaps/",tp,".png",sep='')),width = 8,height = 6,unit = "in",res = 1000)
    print(htmap_list[[tp]])
  dev.off()
  print(htmap)
}

```

# dotplots of genotype effects on cell type proportions


```{r}
dotplot_df = imap_dfr(p_list, function(pmat, tp) {
  cmat = coef_list[[tp]]
  as_tibble(pmat, rownames="genotype") %>%
    pivot_longer(-genotype, names_to="cell_type", values_to="pval") %>%
    left_join(
      as_tibble(cmat, rownames="genotype") %>%
        pivot_longer(-genotype, names_to="cell_type", values_to="coef"),
      by = c("genotype","cell_type")
    ) %>%
    mutate(time_point = tp,
           neglog10_p  = -log10(pval))
})

dotplot_df = dotplot_df %>%
  mutate(time_point = factor(time_point,
                             levels = c("D3","D7","D11","D18"),
                             ordered = TRUE))

```

```{r coef-dotplot-tp,fig.height=9,fig.width=6}




dotplot_list = map(c("D3","D7","D11","D18"), function(tp) {

  
  coef_mtx = coef_list[[tp]]

  genotype_order = rownames(coef_mtx)[ htmap_list[[tp]]$tree_row$order ]
  
  df_tp = filter(dotplot_df, time_point == tp) |>
           mutate(
             genotype  = factor(genotype,
                                levels = genotype_order,
                                ordered = TRUE),
             cell_type = factor(cell_type,
                                levels = celltype_orders,
                                ordered = TRUE)
           ) |>
           drop_na(genotype, cell_type)      # drops cell types not present
  ggplot(df_tp,
         aes(cell_type, genotype,
             size  = neglog10_p,
             color = coef)) +
    geom_point() +
    scale_size(range = c(0.5, 6),
               name  = expression(-log[10]~italic(p))) +
    scale_color_gradient2(low      = "#2166AC",
                          mid      = "white",
                          high     = "#B2182B",
                          midpoint = 0,
                          name     = "Coef") +
    labs(title = tp, x = NULL, y = NULL) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          plot.title  = element_text(hjust = 0.5))
})
names(dotplot_list) = c("D3","D7","D11","D18")
dotplot_list

for(tp in names(dotplot_list)){
  
ggsave(file.path(out_dir, paste0("coefs_dotplots/", tp, ".png",sep='')),
               plot     = dotplot_list[[tp]],
              width = 6,height = 9,
              unit = "in",
               dpi      = 300)
}

pdf(file.path(out_dir,"coefs_dotplots/coefs_dotplots.pdf"), width = 6, height = 9)
for(tp in names(dotplot_list)){
  print(dotplot_list[[tp]])
}
dev.off()
```
```{r coef-dotplot-all,fig.height=6,fig.width=5.5}


# take the order from the heat-map dendrogram of *one* reference TP.
genotype_order = rownames(coef_list[["D18"]])[ htmap_list[["D18"]]$tree_row$order ]

tp_levels      = levels(dotplot_df$time_point)
tp_ct_levels = unlist(
  lapply(tp_levels, \(tp) paste(tp, celltype_orders, sep = "_"))
)

plot_df = dotplot_df %>% 
  mutate(genotype = factor(genotype,
                           levels  = genotype_order,
                           ordered = TRUE),
         tp_ct  = factor(paste(time_point, cell_type, sep = "_"),
                           levels  = tp_ct_levels,
                           ordered = TRUE)) %>% 
  drop_na(genotype, tp_ct)

dotplot = plot_df%>%
  ggplot(aes(x = tp_ct,         
           y = genotype,
           size  = neglog10_p,
           colour = coef)) +
  geom_point() +
  scale_size(range = c(0.3, 4),
             name  = expression(-log[10]~italic(p))) +
  scale_colour_gradient2(low = "#2166AC",
                         mid = "white",
                         high = "#B2182B",
                         midpoint = 0,
                         name = "Coef") +
  labs(x = NULL, y = NULL) +
  theme_bw(base_size = 8) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        panel.grid.major.x = element_blank(),      # optional: declutter grid
        plot.title = element_text(hjust = 0.5))
dotplot
  
ggsave(file.path(out_dir,"coefs_dotplots/all.png"),
               plot  = dotplot,
              width = 5.5,height = 6,
              unit = "in",
               dpi      = 300)

```

# Session information

```{r}
gc()
sessionInfo()
```